---
title: "Untitled"
format: html
---

## Analysis of neutrophil subsets after lipid treatment

```{r}
library(CATALYST)
library(harmony)
library(cowplot)
library(flowCore)
library(scater)
library(SingleCellExperiment)
library(viridis)
library(RColorBrewer)
library(ggforce)
library(ggplot2)
library(ggrepel)
library(ggsci)
library(data.table)
library(emmeans)
library(dplyr)
library(tidyr)
library(ggsignif)
library(rstatix)
library(multcomp)
library(scales)
library(ggthemes)
library(slingshot)
library(ggbeeswarm)
library(ggbreak)
library(DescTools)
library(pak)
library(gghighlight)
library(scattermore)
library(clustree)
library(gridExtra)
library(patchwork)
library(ggbump)
library(stringr)
```

```{r}
setwd("C:/Users/dan94/OneDrive - University College London/UCL_Senior_Research_Fellow/Neutrophil_lipids/GSK_3uM_updated/")
```

## load in FCS files and create flowset object
### list fcs files in folder, choose ones required and then run read.flowset
```{r}
fcs <- list.files(pattern = ".fcs$")
fcs
fs <- read.flowSet(fcs, transformation = F, truncate_max_range = F)
```

## check columns (fluorophores)
## remove any not needed
## when running for the first time, save the columns into an excel to form the basis of the fluorophore metadata
## if this has been created before, just load it in

```{r}
all_cols <- colnames(fs)
all_cols
fs <- fs[,-c(1,2)]
colnames(fs)
cols <- colnames(fs)
cols <- as.data.frame(cols)
#write.csv(cols, file = "Marker.metadata.csv")
markers <- read.csv("Marker.metadata.csv")
markers
```

## use the file names as the basis for the sample metadata
## first add a column to make the sample id (removes excess characters at end of each fcs file name)
## use separate to add individual metadata features
## if script run previously then just load it in
```{r}
names <- data.frame(fcs)
sample_df <- names %>% mutate(sample_id = str_sub(fcs, 1, -24))
sample_df <- separate(sample_df, col = sample_id, into = c("donor", "treatment", "time"), remove = F)
write.csv(sample_df, file = "sample.metadata.csv")
samples <- read.csv("sample.metadata.csv")
```